---
title: "Live coding for teaching computation"
author: "Colin Rundel"
date: "2019-07-10"
output:
  xaringan::moon_reader:
    css: "slides.css"
    seal: FALSE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: monokai-sublime
      countIncrementalSlides: false
      beforeInit: ["https://use.fontawesome.com/releases/v5.9.0/js/all.js"]

---
exclude: true

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
fansi::set_knit_hooks(knitr::knit_hooks)
options(crayon.enabled = TRUE)

knitr::opts_chunk$set(cache = TRUE)

library(ghclass)
library(dplyr)
```

---
class: title_bg

.title[
Live coding
]
.subtitle[
for teaching computation
]


.conference[
.name[
ECME, Dec 2019
]

.bitly[
[bit.ly/rundel_ecme2019](http://bit.ly/rundel_ecme2019)
]

]

.author[
.name[
Colin Rundel
]
.school[
Univ of Edinburgh
]
]

---

## Context

This semester,
* [Math 11176 - Statistical Programming](https://statprog-s1-2019.github.io/)
  * Required (SwDS) MSc course (10 credit)
  * 186 students (109 Stat, 50 OR, 18 ACM, 7 Other)
  
<br/>
  
Previously at Duke,
* [Sta 323 - Statistical Computing](http://www2.stat.duke.edu/~cr173/Sta323_Sp19/)
  * Year 3 UG elective, ~40 students
* [Sta 523 - Statistical Programming](http://www2.stat.duke.edu/~cr173/Sta523_Fa18/)
  * Year 1 MSc required course, ~40 students
* [Sta 790-02 - Advanced Statistical Computing](http://www2.stat.duke.edu/~cr173/Sta790_Sp19/)
  * Year 2 MSc / PhD special topics course, ~15 students
  
---

## Learning Objectives

* Teaching R as a programming language

* Reproducible computing

* Software engineering

* Collaboration

--

<br/><br/>

.center[
*statistics course with programming*

vs

*programming course with statistics*
]

---

## In the classroom

Lectures mostly structured around a traditional format,

* Prepared slides with context on syntax, functions, etc.

* Broken up with in-class exercises (not collected or marked)

  * direct application of the topic(s) just covered
  
  * encouraged to work together and ask quesions
  
  * work through the solution as a class (live coding)

---

## Live Coding

> "the process of designing and implementing a [coding] project in front of class during lecture period" - [Paxton (2002)](https://dl.acm.org/citation.cfm?id=771332)

<br/>

> "Live-coding is an approach to teaching programming by writing actual code during class as part of the lectures. In a live-coding session, the instructor thinks aloud while writing code and the students are able to understand the process of programming by observing the thought processes of the instructor." - [Raj, et al. (2018)](http://pages.cs.wisc.edu/~gerald/papers/LiveCoding.pdf)

---

## Exercise

Using the `flights` data from `nycflights13` answer the following question:

1. How many flights from New York (JFK) to Los Angeles (LAX) did each carrier have in May?

---

## Approach 1 

```{r message=FALSE}
library(dplyr)
library(nycflights13)

flights %>%
  filter(origin == "JFK", dest == "LAX") %>%
  filter(month == 6) %>%
  group_by(carrier, .drop=FALSE) %>%
  summarize(n = n())
```

---
class: center, middle

# Approach 2 - Demo 


---

## Benefits

* Modeling algorithmic thinking 

* Modeling problem solving process

* Modeling debugging processes / finding help

* Interactivity and experimentation

* Comparing alternative solutions

---

## Considerations

Common issue with live coding: 

* students attempting to type everything you do

* they ignore the narrative and miss the context of the problem / solution.

* non-native english speakers can have a harder time

--

Some additional tools for helping with these issues:

* Posting of all code produced (to LMS, GitHub, etc.)

* Recording screencasts (not just lecture capture)

* Broadcasting the code in real time

---

## `livecode`

This an R package, in the early stages of development, for broadcasting a source code file to an  audience in real time.

* Only requires a browser to connect, but also integrates with the RStudio environment

* Removes the need for students to type the code - encourage experimentation via Copy & Paste

* Includes typical IDE features like syntax highlighting

* Also includes line highlighting and messaging functionality

* Some initial trials this semester in Mine's & my course.

---
class: center, middle

# `livecode` demo

---

## Future Directions

Next year Stat Progamming is likely to flip from 2 hrs lecture, 1 hr workshop to 1 hr lecture, 2 hr workshop format

* Adopting a flipped format - record and post lectures, use lecture hours for exerices / live coding
  * Entire git / GitHub component was delivered this way this year
  
* Trial `livecode` more extensively and collect feedback

* Balancing active / passive learning with live coding 

