Homework 1
========================================================

### Task 1 - Implementation



In this function, I used a simple for loop and a series of "if" and "else if" statements.
The "if" statements are broken up into those that check for errors in inputs such as "Inf",
negative values, and non-integers. After working through the possible errors presented, the
second chunk of if statements categorizes the values that are inputted into those divisible
by 3, 5, both, or neither and outputs the corresponding values.

This code was created mainly from my prior knowledge of R as well as class lectures, but I did
seek help from StackOverflow in finding a way to test for integers, linked here: https://stackoverflow.com/questions/3476782/check-if-the-number-is-integer.


```{r}
fizzbuzz = function(v) {
  for (i in v){ #iterates through the values defined in v
    
    #Error cases
    if ((is.integer(i) == FALSE) & (is.double(i) == FALSE)){
      stop("Value is non-numeric")
    }
    else if ((i <= 0) || (i == Inf)){
      stop("Value is not in the acceptable range")
    }
    else if (i%%1 != 0){
      stop("Value is not an integer")
    }
    
    
    #Constructing working cases options
    else if ((i%%3 == 0) & (i%%5 == 0)){
      print("FizzBuzz")
    }
    else if (i%%3 == 0){
      print("Fizz")
    }
    else if (i%%5 == 0){
      print("Buzz")
    }
    else{
      print(i)
    }
  }
}
```

### Task 2 - Validation & Testing

#### Good Inputs

```{r}
fizzbuzz(1)
fizzbuzz(3)
fizzbuzz(5)
fizzbuzz(15)

fizzbuzz(9:15)
fizzbuzz(15:9)
```


#### Bad Inputs

```{r error=TRUE}
fizzbuzz(-1)
fizzbuzz(-3)
fizzbuzz(-5)

fizzbuzz(Inf)
fizzbuzz(-Inf)

fizzbuzz("A")
fizzbuzz(1.5)
fizzbuzz(1i)
fizzbuzz(4i)
```

